Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-06-19 main.tex
Added plots and some tables for tournament results

@@

\begin{abstract}
    We present several powerful strategies for the Iterated
    Prisoner's Dilemma created using reinforcement learning techniques
    (evolutionary and particle swarm algorithms). These strategies are
    trained to perform well against a corpus of over 100 distinct
    opponents, including many well-known strategies from the literature, and all
    the trained strategies win standard tournaments against the total collection
    of other opponents. We also trained variants to win noisy tournaments.
\end{abstract}

@@

@@

% TODO: cite earlier tournaments

@@

The first strategy trained with reinforcement learning in the library is based
on lookup tables. The strategy encodes a set of deterministic responses
based on the opponent's first $n_1$ moves, the opponent's last $m_1$ moves, and
the players last $m_2$ moves. If $n_1 > 0$ then the player has infinite memory
depth, otherwise it has depth $\max{m_1, m_2}$. Although we tried various
combinations of $n_1, m_1,$ and $m_2$, the best performance at the time of
training was obtained for $n_1 = m_1 = m_2 = 2$ and generally for $n_1 > 0$. 
First impressions matter in the IPD. The library includes a strategy
called EvolvedLookerUp2_2_2 which is among the top strategies in the library.

@@

PSO Gambler is a stochastic variant of LookerUp. Instead of deterministically
encoded moves the lookup table emits probabilities which are
used to choose cooperation or defection. The library includes a player trained
with $n_1 = m_1 = m_2 = 2$ that is \emph{mostly deterministic}, with most of the
probabilities being 0 or 1. At one time this strategy outperformed
EvolvedLookerUp2_2_2.

@@

\subsection{ANN}

@@

% TODO: One or more diagrams

@@

We conducted a tournament with all strategies in the Axelrod library. The top
11 performing strategies by mean payoff are all strategies trained to maximize
total payoff against a subset of the strategies (Figure \ref{}). The next strategy is 
Desired Belief Strategy (DBS) which actively analyzes the opponent and responds
accordingly. The next two strategies are Winner12, based on a lookup table, and
Fool Me Once, a strategy that defects indefinitely on the second defection.

Pairwise payoff results are given as a heatmap (Figure \ref{}) which
shows that the many strategies achieve mutual cooperation with each other and
that the top performing strategies never defect first but are able to exploit
weaker strategies that attempt to defect.

The strategies that win the most matches are Defector, Aggravater, followed by
handshaking and zero determinant strategies.

% Table of best strategies
% \includegraphics{strategies_std_boxplot.svg}
% \includegraphics{strategies_std_payoff.svg}
% \includegraphics{strategies_std_winplot.svg}

@@

We also ran noisy tournaments in which there is a 5\% chance that an action
is flipped. The best performing strategies in mean payoff are DBS, designed
to correct for noise, followed by two finite state machine strategies trained
in the presence of noise, and OmegaTFT (also designed to handle noise). The
strategies trained in the presence of noise are also among the best performers
in the absence of noise.

The cluster of mutually cooperative strategies is broken by the noise at 5\%. A
similar collection of players excels at winning matches but again they have
a poor total payoff.

% Table of best strategies
% \includegraphics{strategies_noisy_boxplot.svg}
% \includegraphics{strategies_noisy_payoff.svg}
% \includegraphics{strategies_noisy_winplot.svg}

@@

We trained a variety of strategies using evolutionary algorithms, and in the
case of PSO Gambler particle swarm algorithms. The evolutionary algorithms
used standard techniques, varying strategies by mutation and crossover, and
evaluating the performance against each opponent for many repetitions. The
best performing strategies in each generation are persisted, variants created,
and objective functions computed again. This process continues for approxaimtely
200 generations or until strategies no longer improve significantly.

% Diagram with ANN training performance

@@

The tournament results indicate that pretrained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. A simple evolutionary algorithm produces strategies based on multiple
standard machine learning techniques that are able to achieve a higher mean
score than any other known opponent. All of the trained strategies use many
rounds of the history of play (some using all of it) and outperform memory-one
strategies.

Directly designing an effective strategy is no easy task. Of all the actively
learning strategies in the library, only DBS performs well. The authors trained
a neural network to predict opponents' next moves with high accuracy but basing
a strategy on the predictions proved difficult -- it is necessary to consider
the downstream consequences of a defection and direct training of strategy
outputs generally performed better.

These results suggest that, in opposition to historical tournament results
and community folklore, that complex strategies can be very effective for the
IPD. It is not the complexity of strategies that is disadvantageous, rather it
is apparently too difficult for humans to design complex yet effective
strategies, DBS being a notable exception.

Furthermore, dealing with noise is difficult for most strategies. Two strategies
designed specifically to account for noise, DBS and OmegaTFT, perform well and
only DBS performs better than directly trained finite state machine strategies.

Of all the strategies trained to maximize their mean score all are generally
cooperative, not defecting until the opponent defects. Maximizing for individual
performance across a collection of opponents leads to mutual cooperation despite
the fact that mutual cooperation is an unstable equilibrium for the prisoner's
dilemma. Specifically we note that the reinforcement learning process does not
lead to exploitative zero determinant strategies, which may be a result of the
collection of training strategies, many of which retailate harshly.

Finally, we note that as the library grows, the top performing strategies
sometimes shuffle, and are not retrained regularly. The precise parameters
that are optimal will depend on the pool of opponents. Moreover we have not
extensively trained strategies to determine the minimum parameters that are
sufficient -- neural networks with fewer nodes and features and finite state
machines with fewer states may suffice. See \cite{ashlock2013impact} for
discussion of resource availability for IPD strategies.

@@

@@

>>>>

\usepackage{pdflscape}

@@

\begin{abstract}
    We present tournament results and several powerful strategies for the Iterated
    Prisoner's Dilemma created using reinforcement learning techniques
    (evolutionary and particle swarm algorithms). These strategies are
    trained to perform well against a corpus of over 100 distinct
    opponents, including many well-known strategies from the literature, and all
    the trained strategies win standard tournaments against the total collection
    of other opponents. We also trained variants to win noisy tournaments.
\end{abstract}

@@

% TODO: more

@@

% TODO: discuss earlier tournaments

@@

The first strategy trained with reinforcement learning in the library is based
on lookup tables. The strategy encodes a set of deterministic responses
based on the opponent's first $n_1$ moves, the opponent's last $m_1$ moves, and
the players last $m_2$ moves. If $n_1 > 0$ then the player has infinite memory
depth, otherwise it has depth $\max{m_1, m_2}$. Although we tried various
combinations of $n_1, m_1,$ and $m_2$, the best performance at the time of
training was obtained for $n_1 = m_1 = m_2 = 2$ and generally for $n_1 > 0$. 
First impressions matter in the IPD. The library includes a strategy
called EvolvedLookerUp2\_2\_2 which is among the top strategies in the library.

@@

PSO Gambler is a stochastic variant of LookerUp. Instead of deterministically
encoded moves the lookup table emits probabilities which are
used to choose cooperation or defection. The library includes a player trained
with $n_1 = m_1 = m_2 = 2$ that is \emph{mostly deterministic}, with most of the
probabilities being 0 or 1. At one time this strategy outperformed
EvolvedLookerUp2\_2\_2.

@@

\subsection{ANN: Single Layer Artificial Neural Network}

@@

% TODO: One or more FSM diagrams

@@

We conducted a tournament with all strategies in the Axelrod library. The top
11 performing strategies by mean payoff are all strategies trained to maximize
total payoff against a subset of the strategies (Table \ref{tbl:standard_score}).
The next strategy is
Desired Belief Strategy (DBS) which actively analyzes the opponent and responds
accordingly. The next two strategies are Winner12, based on a lookup table,
Omega TFT, and Fool Me Once, a grudging strategy that defects indefinitely on
the second defection. All strategies in the tournament follow a simple set of
rules in accordance with earlier tournaments:
\begin{itemize}
  \item Players carry no acquired state between matches
  \item Players cannot observe the outcome of other matches
  \item Players cannot identify their opponent by any label or identifier
  \item Players cannot manipulate or inspect their opponents in any way
\end{itemize}

Pairwise payoff results are given as a heatmap (Figure ) which
shows that many strategies achieve mutual cooperation with each other and
that the top performing strategies never defect first but are able to exploit
weaker strategies that attempt to defect.

The strategies that win the most matches are Defector, Aggravater, followed by
handshaking and zero determinant strategies. This includes two handshaking
strategies that were the result of training to maximize Moran process fixation.
No strategies were trained specifically to win matches. None of the top scoring
strategies appear in the top 20 list of strategies ranked by match wins.

% Table of best strategies
\begin{table}[!hbtp]
    \centering
        \input{./tex/standard_scores.tex}
        \caption{Standard Tournament: Top Ranking Strategies by Median Score in 15,000 Tournaments}
        \label{tbl:standard_score}
\end{table}

\begin{landscape}
    \includegraphics[width=\paperwidth]{plots/Standard_boxplot.pdf}
\end{landscape}

% \includegraphics{strategies_std_boxplot.svg}
% \includegraphics{strategies_std_payoff.svg}
% \includegraphics{strategies_std_winplot.svg}

@@

We also ran noisy tournaments in which there is a 5\% chance that an action
is flipped. The best performing strategies in mean payoff are DBS, designed
to correct for noise, followed by two strategies trained
in the presence of noise, Spiteful TFT (TFT but defects indefinitely if the
opponent defects twice consecutively), two strategies trained without noise, then
OmegaTFT (also designed to handle noise).

The strategies trained in the presence of noise are also among the best performers
in the absence of noise. The cluster of mutually cooperative strategies is
broken by the noise at 5\%. A similar collection of players excels at winning
matches but again they have a poor total payoff.

The strategies tallying the most wins are somewhat similar, with Defector, the
handshaking CollectiveStrategy, and Aggravate appearing as the top three again,
followed by Spiteful TFT and strategies designed to be retaliatory. Players ranked
12, 13, and 15 (TF3, Fool Me Once, and Evolved ANN 5) all appear in the top 15
strategies by score, indicating that noise breaks the asymmetry between wins
and scores to some extent.

% Table of best strategies

\begin{table}[!hbtp]
    \centering
        \input{./tex/noisy_scores.tex}
        \caption{Noisy Tournament: Top Ranking Strategies by Median Score in a
        15000 Tournaments with 5\% noise}
        \label{tbl:noisy_score}
\end{table}

\begin{landscape}
    \includegraphics[width=\paperwidth]{plots/Noisy_boxplot.pdf}
\end{landscape}
% \includegraphics{strategies_noisy_payoff.svg}
% \includegraphics{strategies_noisy_winplot.svg}

@@

We trained a variety of strategies using evolutionary algorithms, and in the
case of PSO Gambler particle swarm algorithms. The evolutionary algorithms
used standard techniques, varying strategies by mutation and crossover, and
evaluating the performance against each opponent for many repetitions. The
best performing strategies in each generation are persisted, variants created,
and objective functions computed again. This process continues for approximately
200 generations or until strategies no longer improve significantly.

% Todo: Diagram with ANN training performance?

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. A simple evolutionary algorithm produces strategies based on multiple
standard machine learning techniques that are able to achieve a higher mean
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies (though the trained memory one strategy performs
well). The generic structure of the trained strategy did not appear to be
critical -- strategies based on lookup tables, finite state machines, and stochastic
variants all performed well. Single layer neural networks also performed well
though these had some aspect of human involvement in the selection of features.
The success of the other strategy types suggests that a deeper network that
incorporates feature engineering would likely also perform well.

In opposition to historical tournament results and community folklore,
our results show that complex strategies can be very effective for the
IPD. It is not the complexity of strategies that is disadvantageous; rather that directly
designing a broadly effective strategy is no easy task. Of all the human-designed
strategies in the library, only DBS consistently performs well, and it is
substantially more complex than traditional tournament winners like TFT, OmegaTFT,
and zero determinant strategies. Furthermore, dealing with noise is difficult
for most strategies. Two strategies designed specifically to account for noise,
DBS and OmegaTFT, perform well and only DBS performs better than our trained
strategies.

Of the strategies trained to maximize their mean score all are generally
cooperative, not defecting until the opponent defects. Maximizing for individual
performance across a collection of opponents leads to mutual cooperation despite
the fact that mutual cooperation is an unstable equilibrium for the prisoner's
dilemma. Specifically we note that the reinforcement learning process for maximizing
payout does not
lead to exploitative zero determinant strategies, which may be a result of the
collection of training strategies, many of which retaliate harshly.

Finally, we note that as the library grows, the top performing strategies
sometimes shuffle, and are not retrained regularly. Most of the strategies were
trained on an earlier version of the library (v2.2.0) that did not include DBS
and several other opponents. The precise parameters that are optimal will depend
on the pool of opponents. Moreover we have not extensively trained strategies to
determine the minimum parameters that are sufficient -- neural networks with
fewer nodes and features and finite state machines with fewer states may suffice.
See \cite{ashlock2013impact} for discussion of resource availability for IPD strategies.

@@

% TODO: Full strategy list

@@

\end{document}
